<<<<<<< HEAD
---
title: "Final Ovarian Cancer Project Script"
output: html_notebook
---


=======
>>>>>>> 947b3ba59bf8677cda0739aa85bbd1f2619550e1
```{r}
if (!require("BiocManager")) install.packages("BiocManager")

if (!require("TCGAbiolinks")) BiocManager::install("TCGAbiolinks")
library(TCGAbiolinks)

if (!require("survival")) install.packages("survival")
library(survival)

if (!require("survminer")) install.packages("survminer")
library(survminer)

if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

if (!require("maftools")) BiocManager::install("maftools")
library(maftools)

if (!require("SummarizedExperiment")) BiocManager::install("SummarizedExperiment")
library(SummarizedExperiment)

if (!require("DESeq2")) BiocManager::install("DESeq2")
library(DESeq2)

if (!require("sesame")) BiocManager::install("sesame")
library(sesame)

if (!require("sesameData")) BiocManager::install("sesameData")
library(sesameData)

if (!require("limma")) BiocManager::install("limma")
library(limma)


output_dir <- "qbio490_fa24_final_ov/outputs"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
setwd(output_dir) 

#Query, download, and prepare RNA-seq data
rna_query <- GDCquery(
  project = "TCGA-OV",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
<<<<<<< HEAD
GDCdownload(rna_query)  #Download the RNA-seq data
rna_se <- GDCprepare(rna_query)  #Prepare the RNA-seq data
=======
#GDCdownload(rna_query)  # Download the data
rna_se <- GDCprepare(rna_query)  # Prepare the data into SummarizedExperiment format
ov_rna_clinical <- as.data.frame(rna_se@colData)  # Extract clinical metadata from RNA-seq
```
>>>>>>> 947b3ba59bf8677cda0739aa85bbd1f2619550e1

#Query, download, and prepare clinical data
clin_query <- GDCquery(
  project = "TCGA-OV",
  data.category = "Clinical",
  data.type = "Clinical Supplement",
  data.format = "BCR Biotab"
)
<<<<<<< HEAD
GDCdownload(clin_query)  #Download the clinical data
clinical_data <- GDCprepare(clin_query)  #Clinical data prep
ov_clinical <- clinical_data$clinical_patient_ov[-c(1, 2), ]  #Excludes non-patient rows
colnames(ov_clinical)[colnames(ov_clinical) == "bcr_patient_barcode"] <- "Tumor_Sample_Barcode"
=======
#GDCdownload(clin_query)  # Download the data
clinical.BCRtab.all <- GDCprepare(clin_query)  # Prepare the data
ov_clinical <- clinical.BCRtab.all$clinical_patient_ov[-c(1,2),]  # Remove non-patient data rows
colnames(ov_clinical)[colnames(ov_clinical) == "bcr_patient_barcode"] <- "Tumor_Sample_Barcode"  # Rename key column
```
>>>>>>> 947b3ba59bf8677cda0739aa85bbd1f2619550e1

#Query, download, and prepare mutation data
maf_query <- GDCquery(
  project = "TCGA-OV",
  data.category = "Simple Nucleotide Variation",
  access = "open",
  data.type = "Masked Somatic Mutation",
  workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)
<<<<<<< HEAD
GDCdownload(maf_query)  #mutation data
maf <- GDCprepare(maf_query)  #mutation data prep
maf_object <- read.maf(maf = maf, clinicalData = ov_clinical, isTCGA = TRUE)
=======
#GDCdownload(maf_query)  # Download the mutation data
maf <- GDCprepare(maf_query)  # Prepare the mutation data
maf_object <- read.maf(maf = maf, clinicalData = ov_clinical, isTCGA = TRUE)  # Create a MAF object

# Add BRCA mutation statuses to clinical data
brca1_mask <- maf_object@data$Hugo_Symbol == "BRCA1"
brca2_mask <- maf_object@data$Hugo_Symbol == "BRCA2"

brca1_barcodes <- unique(maf_object@data$Tumor_Sample_Barcode[brca1_mask])
brca2_barcodes <- unique(maf_object@data$Tumor_Sample_Barcode[brca2_mask])
>>>>>>> 947b3ba59bf8677cda0739aa85bbd1f2619550e1

#Adding BRCA mutation statuses to clinical data
brca1_barcodes <- unique(maf_object@data$Tumor_Sample_Barcode[maf_object@data$Hugo_Symbol == "BRCA1"])
brca2_barcodes <- unique(maf_object@data$Tumor_Sample_Barcode[maf_object@data$Hugo_Symbol == "BRCA2"])
ov_clinical$BRCA1 <- ifelse(ov_clinical$Tumor_Sample_Barcode %in% brca1_barcodes, "Mutated", "Not Mutated")
ov_clinical$BRCA2 <- ifelse(ov_clinical$Tumor_Sample_Barcode %in% brca2_barcodes, "Mutated", "Not Mutated")
ov_clinical$BRCA <- ifelse(ov_clinical$BRCA1 == "Mutated" | ov_clinical$BRCA2 == "Mutated", "Mutated", "Not Mutated")
ov_clinical$BRCA <- factor(ov_clinical$BRCA, levels = c("Not Mutated", "Mutated"))

#Survival data prep
ov_clinical$death_days_to <- suppressWarnings(as.numeric(ov_clinical$death_days_to))
ov_clinical$last_contact_days_to <- suppressWarnings(as.numeric(ov_clinical$last_contact_days_to))
ov_clinical$survival_time <- ifelse(
  !is.na(ov_clinical$death_days_to),
  ov_clinical$death_days_to,
  ov_clinical$last_contact_days_to
)
ov_clinical <- ov_clinical[!is.na(ov_clinical$survival_time), ]
ov_clinical$death_event <- ifelse(ov_clinical$vital_status == "Dead", TRUE, FALSE)

<<<<<<< HEAD
#Kaplan-Meier survival analysis
=======

# KM plots
```{r}
# Create Kaplan-Meier survival curves for BRCA mutation groups
>>>>>>> 947b3ba59bf8677cda0739aa85bbd1f2619550e1
survival_object <- Surv(time = ov_clinical$survival_time, event = ov_clinical$death_event)
fit_brca <- survfit(survival_object ~ BRCA, data = ov_clinical)  # Combined BRCA mutations
fit_brca1 <- survfit(survival_object ~ BRCA1, data = ov_clinical)  # BRCA1 mutations
fit_brca2 <- survfit(survival_object ~ BRCA2, data = ov_clinical)  # BRCA2 mutations

#Create KM plot for BRCA mutation status
plot_brca <- ggsurvplot(
  fit_brca,
  pval = TRUE,
  risk.table = TRUE,
  title = "Kaplan-Meier Curve: BRCA Mutation Status",
  xlab = "Time (days)",
  ylab = "Survival Probability",
  legend.labs = c("Not Mutated", "Mutated")
)

plot_brca1 <- ggsurvplot(
  fit_brca1,
  pval = TRUE,
  risk.table = TRUE,
  title = "Kaplan-Meier Curve: BRCA1 Mutation Status",
  xlab = "Time (days)",
  ylab = "Survival Probability",
  legend.labs = c("Not Mutated", "Mutated")
)

plot_brca2 <- ggsurvplot(
  fit_brca2,
  pval = TRUE,
  risk.table = TRUE,
  title = "Kaplan-Meier Curve: BRCA2 Mutation Status",
  xlab = "Time (days)",
  ylab = "Survival Probability",
  legend.labs = c("Not Mutated", "Mutated")
)


#Save KM plot
ggsave("KM_BRCAMutations.png", plot = plot_brca$plot, width = 8, height = 6)
ggsave("KM_BRCA1.png", plot = plot_brca1$plot, width = 8, height = 6)
ggsave("KM_BRCA2.png", plot = plot_brca2$plot, width = 8, height = 6)


#Differential Expression Analysis
rna_counts <- assay(rna_se, "unstranded")
rna_clinical <- as.data.frame(rna_se@colData)
rna_clinical$BRCA <- ov_clinical$BRCA[match(substr(rownames(rna_clinical), 1, 12), ov_clinical$Tumor_Sample_Barcode)]
rna_clinical <- rna_clinical[!is.na(rna_clinical$BRCA), ]
rna_counts <- rna_counts[, colnames(rna_counts) %in% rownames(rna_clinical)]

dds <- DESeqDataSetFromMatrix(
  countData = rna_counts,
  colData = rna_clinical,
  design = ~ BRCA
)
dds <- DESeq(dds)
results <- as.data.frame(results(dds, contrast = c("BRCA", "Mutated", "Not Mutated")))

#Save differential expression results
write.csv(results, "DEG_results.csv", row.names = TRUE)

#Generate Volcano Plot
results$logP <- -log10(results$padj)
results$Significance <- ifelse(
  results$padj < 0.05 & abs(results$log2FoldChange) > 1,
  "Significant",
  "Not Significant"
)
volcano_plot <- ggplot(results, aes(x = log2FoldChange, y = logP, color = Significance)) +
  geom_point(alpha = 0.8, size = 1.5) +
  theme_minimal() +
  labs(
    title = "Volcano Plot: Differential Expression Analysis",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value"
  )
ggsave("VolcanoPlot.png", plot = volcano_plot, width = 8, height = 6)

#Generate Lollipop Plots
brca1_maf <- subsetMaf(maf = maf_object, tsb = brca1_barcodes)
brca2_maf <- subsetMaf(maf = maf_object, tsb = brca2_barcodes)
pdf(file = "Lollipop_BRCA1.pdf", width = 8, height = 6)
lollipopPlot(maf = brca1_maf, gene = "BRCA1")
dev.off()

# Generate and save lollipop plot for BRCA2
pdf(file = "Lollipop_BRCA2.pdf", width = 8, height = 6)
lollipopPlot(maf = brca2_maf, gene = "BRCA2")
dev.off()

#Output top 100 DEGs
significant_genes <- results[results$padj < 0.05 & !is.na(results$padj), ]
top_100_overexpressed <- head(significant_genes[order(-significant_genes$log2FoldChange), ], 100)
top_100_underexpressed <- head(significant_genes[order(significant_genes$log2FoldChange), ], 100)
write.csv(top_100_overexpressed, "Top100_Overexpressed.csv")
write.csv(top_100_underexpressed, "Top100_Underexpressed.csv")

```
<<<<<<< HEAD
=======

# Prepare RNA-seq count data for DESeq2
```{r}
# Extract RNA-seq count matrix
rna_counts <- assay(rna_se, "unstranded")  # Replace with the correct assay (e.g., stranded_first) if needed
rna_counts <- as.data.frame(rna_counts)

# Match clinical data with RNA-seq samples
rna_clinical <- as.data.frame(rna_se@colData)
rna_clinical$BRCA <- ov_clinical$BRCA[match(substr(rownames(rna_clinical), 1, 12), ov_clinical$Tumor_Sample_Barcode)]

# Filter out samples with missing BRCA mutation status
rna_clinical <- rna_clinical[!is.na(rna_clinical$BRCA), ]

# Subset the count data to include only samples with BRCA mutation status
rna_counts <- rna_counts[, colnames(rna_counts) %in% rownames(rna_clinical)]

# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = rna_counts,
  colData = rna_clinical,
  design = ~ BRCA  # Model BRCA mutation status
)

# Run DESeq2 pipeline
dds <- DESeq(dds)

# Extract results
results <- results(dds, contrast = c("BRCA", "Mutated", "Not Mutated"))
results <- as.data.frame(results)

# Save DEG results to a file
output_dir <- "outputs"
if (!dir.exists(output_dir)) dir.create(output_dir)
write.csv(results, file = paste0(output_dir, "/DEG_results.csv"), row.names = TRUE)

# Filter significant genes (adjusted p-value < 0.05)
significant_genes <- results[results$padj < 0.05, ]
write.csv(significant_genes, file = paste0(output_dir, "/Significant_DEG.csv"), row.names = TRUE)
```


```{r}
# Remove rows with NA values in log2FoldChange or padj
filtered_results <- results[!is.na(results$log2FoldChange) & !is.na(results$padj), ]

# Recalculate log10 adjusted p-value
filtered_results$logP <- -log10(filtered_results$padj)
```


```{r}
# Ensure log2FoldChange and logP are finite
filtered_results <- filtered_results[is.finite(filtered_results$log2FoldChange) & is.finite(filtered_results$logP), ]
```


```{r}
# Add a new column for negative log10 of adjusted p-value
filtered_results$logP <- -log10(filtered_results$padj)
```

# Convert ENSEMBL ID to Gene Label
```{r}
filtered_results$ensembl_id <- gsub("\\..*$", "", rownames(filtered_results))  # Remove version numbers if present
# Load biomaRt
if (!require("biomaRt")) BiocManager::install("biomaRt")
library(biomaRt)

# Connect to the Ensembl database
ensembl <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", mirror = "useast")

# Get gene names for the Ensembl IDs
gene_mapping <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  filters = "ensembl_gene_id",
  values = filtered_results$ensembl_id,
  mart = ensembl
)

# Merge the gene names back into the filtered_results
filtered_results <- merge(
  filtered_results, 
  gene_mapping, 
  by.x = "ensembl_id", 
  by.y = "ensembl_gene_id", 
  all.x = TRUE
)

# Rename the column for better readability
colnames(filtered_results)[colnames(filtered_results) == "external_gene_name"] <- "Gene_Name"

```

# volcano w/ label and adjusted threshold
```{r}
# Install and load ggrepel if not already installed
if (!require("ggrepel")) install.packages("ggrepel")
library(ggrepel)

# Add a new column for negative log10 of adjusted p-value
filtered_results$logP <- -log10(filtered_results$padj)

# Define significance with separate categories for overexpressed and underexpressed genes
filtered_results$Significance <- ifelse(
  filtered_results$padj < 0.05 & filtered_results$log2FoldChange > 2, "Overexpressed",
  ifelse(
    filtered_results$padj < 0.05 & filtered_results$log2FoldChange < -2, "Underexpressed",
    "Not Significant"
  )
)

# Add gene names to significant points for labeling
filtered_results$GeneLabel <- ifelse(
  filtered_results$Significance %in% c("Overexpressed", "Underexpressed"), 
  filtered_results$Gene_Name, 
  NA
)

# Generate the volcano plot with different colors for over- and underexpressed genes
volcano_plot <- ggplot(filtered_results, aes(x = log2FoldChange, y = logP, color = Significance)) +
  geom_point(alpha = 0.8, size = 1.5) +
  geom_text_repel(
    aes(label = GeneLabel),
    size = 3,
    max.overlaps = 8,  # Adjust this to limit the number of overlapping labels
    box.padding = 0.3,
    point.padding = 0.3
  ) +
  scale_color_manual(values = c(
    "Not Significant" = "grey",
    "Overexpressed" = "#042f50",
    "Underexpressed" = "#3E8F99"
  )) +
  theme_minimal() +
  labs(
    title = "Differential Gene Expression: BRCA1/2-Mutated vs. Wild-Type Patients",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value"
  )

# Define output directory
output_dir <- "outputs"
if (!dir.exists(output_dir)) dir.create(output_dir)

# Save volcano plot to the output directory
ggsave(filename = paste0(output_dir, "/VolcanoPlot_OverUnderDifferentiated.png"), plot = volcano_plot, width = 8, height = 6)

# Display the volcano plot
print(volcano_plot)


```

# Output top 10 DEGs
```{r}
# Filter significant genes (optional, based on adjusted p-value)
significant_genes <- filtered_results[filtered_results$padj < 0.05, ]

# Sort genes by log2FoldChange to identify overexpressed and underexpressed DEGs
overexpressed_genes <- significant_genes[order(-significant_genes$log2FoldChange), ]  # Sort in descending order
underexpressed_genes <- significant_genes[order(significant_genes$log2FoldChange), ]  # Sort in ascending order

# Extract top 10 overexpressed and underexpressed genes
top_10_overexpressed <- head(overexpressed_genes, 10)
top_10_underexpressed <- head(underexpressed_genes, 10)

# Print the top 10 overexpressed and underexpressed genes
print("Top 10 Overexpressed Genes:")
print(top_10_overexpressed)

print("Top 10 Underexpressed Genes:")
print(top_10_underexpressed)

# Save results to CSV files
write.csv(top_10_overexpressed, file = "top_10_overexpressed_genes.csv", row.names = FALSE)
write.csv(top_10_underexpressed, file = "top_10_underexpressed_genes.csv", row.names = FALSE)

```

# Output top 100 DEGs
```{r}
# Filter significant genes (optional, based on adjusted p-value)
significant_genes <- filtered_results[filtered_results$padj < 0.05, ]

# Sort genes by log2FoldChange to identify overexpressed and underexpressed DEGs
overexpressed_genes <- significant_genes[order(-significant_genes$log2FoldChange), ]  # Sort in descending order
underexpressed_genes <- significant_genes[order(significant_genes$log2FoldChange), ]  # Sort in ascending order

# Extract top 10 overexpressed and underexpressed genes
top_100_overexpressed <- head(overexpressed_genes, 100)
top_100_underexpressed <- head(underexpressed_genes, 100)

# Print the top 10 overexpressed and underexpressed genes
print("Top 100 Overexpressed Genes:")
print(top_100_overexpressed)

print("Top 100 Underexpressed Genes:")
print(top_100_underexpressed)

# Save results to CSV files
write.csv(top_100_overexpressed, file = "top_100_overexpressed_genes.csv", row.names = TRUE)
write.csv(top_100_underexpressed, file = "top_100_underexpressed_genes.csv", row.names = TRUE)

```

# lollipop plots
```{r}
# obtain the mutation clinical data frame
maf_clinical <- maf_object@clinical.data

# create a column in the mutation clinical data frame that checks to see if patients are in the list of BRCA1/2 mutant barcodes; if they are not present, put "None"
maf_clinical$BRCA_type <- ifelse(maf_clinical$Tumor_Sample_Barcode %in% brca1_barcodes, "BRCA1", ifelse(maf_clinical$Tumor_Sample_Barcode %in% brca2_barcodes, "BRCA2", "None"))

# factorize column so it can be used in analysis
maf_clinical$BRCA_type <- factor(maf_clinical$BRCA_type)

# filter out the BRCA1 patients, gather their barcodes, and subset the maf by those barcodes
brca1_filter <- ifelse(maf_clinical$BRCA_type == 'BRCA1', TRUE, FALSE)
brca1_patient_barcodes <- maf_clinical$Tumor_Sample_Barcode[brca1_filter]
brca1_maf <- subsetMaf(maf = maf_object,
                       tsb = brca1_patient_barcodes)

# filter out the BRCA2 patients, gather their barcodes, and subset the maf by those barcodes
brca2_filter <- ifelse(maf_clinical$BRCA_type == 'BRCA2', TRUE, FALSE)
brca2_patient_barcodes <- maf_clinical$Tumor_Sample_Barcode[brca2_filter]
brca2_maf <- subsetMaf(maf = maf_object, tsb = brca2_patient_barcodes)

# create lollipop plots using the subsetted BRCA1/2 maf dataframes
lollipopPlot(maf = brca1_maf, gene = 'BRCA1')

lollipopPlot(maf = brca2_maf, gene = 'BRCA2')

```
>>>>>>> 947b3ba59bf8677cda0739aa85bbd1f2619550e1
